The Design, Semantics, and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-10-09 paper.tex
Secao VM

\usepackage{listings}
%\usepackage{textcomp}
\usepackage{url}
\lstset{
%columns=fullflexible,
%basicstyle=\ttfamily,
escapeinside={||},
    %mathescape=true,
    language=C, % choose the language of the code
    basicstyle=\fontfamily{pcr}\selectfont\scriptsize\color{black},
    keywordstyle=\color{black}\bfseries, % style for keywords
    numbers=none, % where to put the line-numbers
    numberstyle=\tiny, % the size of the fonts that are used for the line-numbers
    backgroundcolor=\color{light},
    showspaces=false, % show spaces adding particular underscores
    showstringspaces=false, % underline spaces within strings
    showtabs=false, % show tabs within strings adding particular underscores
    %frame=single, % adds a frame around the code
    tabsize=2, % sets default tabsize to 2 spaces
    %rulesepcolor=\color{gray}
    captionpos=b, % sets the caption-position to bottom
    breaklines=false, % sets automatic line breaking
    %breakatwhitespace=false,
    numbersep=2em,
    % C was used in the blocksworld example to refer to block C and nowhere else
    emph={par,or,hor,do,end,loop,await,emit,input,event,call,with,%
          var,and,then,else,return,pure,deterministic,nohold,finalize,%
          class, every, FOREVER, this, spawn, in, pool, watching, until, 
          interface, each, abort, when, signal, PROC, CHAN, SIGNAL, PAR, not,
          bool, data, tag, escape, new, traverse},
    emphstyle={\bfseries},
    commentstyle=\color{dark}\scriptsize,
    %xleftmargin=20pt,
    %xrightmargin=20pt,
    framesep=20pt,
    %upquote=true,
    %aboveskip={1.5\baselineskip},
}

@@

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
input int A, B, C;
do
    var int a = await A;
end
do
    var int b = await B;
end
par/and do
    await B;
with
    await C;
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
union {             // sequence
    int a_1;        //   do_1
    int b_2;        //   do_2
    struct {        //   par/and
        int _and_3: 1;
        int _and_4: 1;
    };
} MEM ;
\end{lstlisting}
\end{minipage}
%\rule{14cm}{0.37pt}
\caption{
A program with blocks in sequence and in parallel, with corresponding memory 
layout that the compiler generates.
{\small %\textmd{
}%}
\label{lst.impl.mem}
}
\end{figure}

@@

#ifdef EVT_PHOTO_READDONE
    event void Photo.readDone (uint16_t val) {
        ceu_sys_go(EVT_PHOTO_READDONE, &val);
    }
#endif

@@

    <...>   // other events
}
\end{lstlisting}
%\rule{14cm}{0.37pt}
\caption{
%{\small %\textmd{
The \emph{TinyOS} binding for \CEU.
%}%}
\label{lst.impl.tinyos}
}
\end{figure}

@@

The communication between scripts in \CEU and the components in the \VM is 
mostly through events:
scripts \code{emit} requests through \code{output} events and \code{await} 
answers through \code{input} events.
%
Terra also provides system calls for initialization and configuration of 
components.
%
TODO: explicar Figura 21.
\begin{comment}
Figure~\ref{lst.terra.defs} illustrates a program interface declaration in \CEU
(in the left), and associated binding code in \C

 with respective binding customization scenario.
This customization sample uses a small data structure, two output events, one input event, and two functions.
The left listing is the definition file to be included into the \CEU script program and the right listing is the
definition file to be included into the VM custom module implementation.
\end{comment}

\begin{figure}
\begin{minipage}[t]{0.46\linewidth}
\begin{lstlisting}[morekeywords={ubyte,ushort,output,function,regtype}]
// Message definition
regtype sensorMsg with
    var ubyte  id;
    var ushort value;
end

// Output events definition
output void REQ_TEMP void          1;
output void SEND_SENSOR sensorMsg  2;

// Input events definition
input void TEMP void               1;

// functions definitions
function ubyte getNodeId()         1;
function ubyte queuePut(sensorMsg) 2;
.
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.51\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em,
morekeywords={nx_struct,nx_uint8_t,nx_uint16_t,uint8_t,uint32_t}]

typedef nx_struct sensorMsg{
  nx_uint8_t  id;
  nx_uint16_t value;
} sensorMsg_t;

enum {
  O_REQ_TEMP    = 1;
  O_SEND_SENSOR = 2;

  I_TEMP        = 1;

  F_GETNODEID   = 1;
  F_QUEUEPUT    = 2;
};

\end{lstlisting}
\end{minipage}
\caption{
%{\small %\textmd{
Terra control block sample and the equivalent definition for VM implementation.
%}%}
\label{lst.terra.defs}
}
\end{figure}

TODO: dizer o que eh manual/automatico nos lsts de ceu

TODO-VM: exemplo de implementacao/uso de um componente
- Terra: binding em nesC
- Ceu: header de configuracao
- (analogo a Figura 19)

Considering the compilation process illustrated in Figure~\ref{fig.impl} for 
the \C back end, the main difference resides in the \emph{Code Generation} 
phase, which here outputs assembly instructions for the Terra \VM, instead of 
\C.

TODO-VM: falar um pouco do assembly/bytecode, quais as decisoes envolvidas
- (analogo a Figura 18)

During compilation process the script is broken in trails in the same way it is 
done in the \CEU compiler explained in Section 4.
The Terra virtual machine implements a similar \CEU scheduler to execute the 
script trails, but the VM
uses the real trails address instead of labels. In this way is possible to jump directly to each script
partitions without using a dispatcher.
The Terra VM instruction set includes operations to support the \CEU execution 
model.
Basically, it is composed by instructions to define a pending trail (\code{tkins}),
a pending event(using \code{set\_c} to set a value in the memory),
a check operation to verify the \code{par/and} termination(\code{chkret}),
a pending timer event (\code{clken}), and some others like to clear trails data or suspend execution.
All pending entries are managed by the scheduler, that also responds for timer and input events.
All trails are executed as a TinyOS task, this ensure that each trail runs to completion avoiding race condition between then.

@@

\begin{comment}
%-------------------------------------------------
Figure~\ref{lst.impl.custom} presents a partial implementation for the elements 
introduced in the Figure~\ref{lst.impl.defs}.
The two first functions are called from the VM decoder, the fist function \code{procOutevt()} dispatch any defined output events
and the function \code{callFunction()} dispatch any defined customized functions.
Lines 17--22 shows an example of a custom function that returns a value via stack.
Lines 24--27 has a example of an external event call.
Lines 29--34 presents an example how a input event is queued to the VM engine.
\end{comment}

@@

\begin{figure}
\begin{minipage}[t]{0.81\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em,
morekeywords={uint8_t,uint16_t,uint32_t,error_t,command}]
// Output event dispatcher
command void VM.procOutEvt(uint8_t id, uint32_t value){
  switch (id){
    case O_REQ_TEMP    : proc_req_temp(id,value);    break;
    case O_SEND_SENSOR : proc_send_sensor(id,value); break;
  }
}

// Function dispatcher
command void VM.callFunction(uint8_t id){
  switch (id){
    case F_GETNODEID : func_getNodeId(id); break;
    case F_QUEUEPUT  : func_queuePut(id);  break;
  }
}

// Pushing a value to the stack
void  func_getNodeId(uint16_t id){
  uint16_t stat;
  stat = TOS_NODE_ID;
  signal VM.push(stat);
}

// Calling a output event
void  proc_req_temp(uint16_t id, uint32_t value){
  call S_TEMP.read();
}

// Queueing an input event + value
uint16_t lastTemp;
event void S_TEMP.readDone(error_t result, uint16_t val)
  lastTemp = val;
  signal VM.queueEvt(I_TEMP, 0, &lastTemp);
}

\end{lstlisting}
\end{minipage}

\caption{
%{\small %\textmd{
VM Customization -- input/output events and functions.
%}%}
\label{lst.impl.custom}
}
\end{figure}

@@

\emph{(As subsecoes serao retiradas do texto ao final...)}

\subsection{Semantics}

@@

A number of synchronous languages have been designed to interoperate with C, 
such as \emph{Reactive C}~\cite{rp.rc}, Protothreads~\cite{wsn.protothreads}, 
\emph{PRET-C}~\cite{rp.pretc} and \emph{SC}~\cite{rp.synchc}.
%
They offer Esterel-like parallel compositions with communication via shared 
variables, relying on deterministic scheduling to preserve determinism.
%
However, it is the responsibility of the programmer to specify the execution 
order for threads, based on either explicit priorities, or source code lexical 
order.
%
These languages have a tick-based notion of time similar to Esterel, which 
prevents the event-based temporal analysis of \CEU.

@@

TODO-VM

Plummer et al.~\cite{esterel.vm} present a reactive virtual machine for 
embedded systems.
This VM uses the Esterel programming language as scripting language and implements
a specific instruction set to the Esterel concurrent and multi-thread semantic.
They focus in code reduction instead of on efficiency.
Experimentally, they found the bytecode was about half the size of optimized
native assembly code generated from C.
The Terra virtual machine also implements a customized instruction set for a reactive
programming model, but it uses the \CEU model for concurrence and event 
control.
The Terra component based approach also contributes to have a very small bytecode size.
In our experiments, depending of the abstraction level of the pre-built customized components,
the bytecode size reduction may reach about 99\%.

\begin{comment}
More recently, Wireless Sensor Networks (WSNs) emerged as an active research 
area for highly constrained embedded concurrency, resulting in the development 
of many synchronous languages~\cite{wsn.protothreads,wsn.sol,wsn.osm}.
%
offer lightweight cooperative multi-threading for embedded systems.
Its stackless implementation reduces memory consumption but precludes support 
for local variables.
\CEU also avoids the use of stacks for trails, but preserves support for locals 
by calculating the required memory at compile time.
%
SOL~\cite{wsn.sol} and OSM~\cite{wsn.osm} provide parallel state machines for 
WSNs, offering a formal and mature model for programming embedded systems.
However, the main contributions of \CEU, stacked execution for internal 
events and safe support for shared-memory concurrency, do not directly adapt to 
the state-machine formalism.

In common among the referred works is the agreement in providing low-level 
access (e.g., systems calls and shared-memory) and lock-free concurrency that 
precludes race conditions on programs.
However, they do not specify an execution order for tasks reacting to the same 
external stimulus~\cite{esterel.primer,wsn.protothreads,wsn.osm}.
This way, if two tasks access the same resource concurrently, even if the 
accesses are race free, the final outcome is nondeterministic.
As discussed in Section~\ref{sec.safety.det}, \CEU refuses programs with such 
behavior.

On the opposite side of the spectrum of concurrency models, asynchronous 
languages for embedded systems~\cite{wsn.mantisos,arduino.occam}
assume time independence among processes and are more appropriate for 
applications with a low synchronization rate or for those involving
algorithmic-intensive problems.
\end{comment}

\subsection{Implementation}

@@

TODO-VM

>>>>

\usepackage{listings}
%\usepackage{textcomp}
\usepackage{url}
\lstset{
%columns=fullflexible,
%basicstyle=\ttfamily,
escapeinside={||},
    %mathescape=true,
    language=C, % choose the language of the code
    basicstyle=\fontfamily{pcr}\selectfont\scriptsize\color{black},
    keywordstyle=\color{black}\bfseries, % style for keywords
    numbers=none, % where to put the line-numbers
    numberstyle=\tiny, % the size of the fonts that are used for the line-numbers
    backgroundcolor=\color{light},
    showspaces=false, % show spaces adding particular underscores
    showstringspaces=false, % underline spaces within strings
    showtabs=false, % show tabs within strings adding particular underscores
    %frame=single, % adds a frame around the code
    tabsize=2, % sets default tabsize to 2 spaces
    %rulesepcolor=\color{gray}
    captionpos=b, % sets the caption-position to bottom
    breaklines=false, % sets automatic line breaking
    %breakatwhitespace=false,
    numbersep=2em,
    % C was used in the blocksworld example to refer to block C and nowhere else
    emph={par,or,hor,do,end,loop,await,emit,input,event,call,with,%
          var,and,then,else,return,pure,deterministic,nohold,finalize,%
          class, every, FOREVER, this, spawn, in, pool, watching, until, 
          interface, each, abort, when, signal, PROC, CHAN, SIGNAL, PAR, not,
          bool, data, tag, escape, new, traverse,implementation,output},
    emphstyle={\bfseries},
    commentstyle=\color{dark}\scriptsize,
    %xleftmargin=20pt,
    %xrightmargin=20pt,
    framesep=20pt,
    %upquote=true,
    %aboveskip={1.5\baselineskip},
}

@@

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
input int A, B, C;
do
    var int a = await A;
end
do
    var int b = await B;
end
par/and do
    await B;
with
    await C;
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
union {             // sequence
    int a_1;        //   do_1
    int b_2;        //   do_2
    struct {        //   par/and
        int _and_3: 1;
        int _and_4: 1;
    };
} MEM ;
\end{lstlisting}
\end{minipage}
%\rule{14cm}{0.37pt}
\caption{
A program with blocks in sequence and in parallel, with corresponding memory 
layout generated by the compiler.
{\small %\textmd{
}%}
\label{lst.impl.mem}
}
\end{figure}

@@

#ifdef EVT_PHOTO_READDONE
    event void Photo.readDone (int val) {
        ceu_sys_go(EVT_PHOTO_READDONE, &val);
    }
#endif

@@

    <...>   // other events
}
\end{lstlisting}
%\rule{14cm}{0.37pt}
\caption{
%{\small %\textmd{
The \emph{TinyOS} binding for \CEU.
This platform-dependent template includes the \C files generated from the 
original application in \CEU (\code{ceu\_app.h} and \code{ceu\_app.c}) for the 
\emph{Final generation} phase of Figure~\ref{fig.impl}.
%}%}
\label{lst.impl.tinyos}
}
\end{figure}

@@

The communication between scripts in \CEU and the components in the \VM is 
mostly through events:
scripts \code{emit} requests through \code{output} events and \code{await} 
answers through \code{input} events.
%
Terra also provides system calls for initialization and configuration of 
components (e.g., \emph{getters} and \emph{setters}).
%
The left of Figure~\ref{lst.terra.defs} shows a \CEU interface with the 
available functionality for a customized \VM with temperature and radio 
components.
The right of the figure shows the associated binding for output events (ln.  
1--8), input events (ln. 10--14), and system calls (ln. 16--22).
%
Note that all applications for the customized \VM must comply with the same 
interface.
In contrast, the template-based \C back end allows applications to choose all 
possible combinations of functionalities from the underlying platform at 
compile time (as illustrated in Figure~\ref{lst.impl.tinyos}).

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[morekeywords={function}]
// Output events
output void REQUEST_TEMPERATURE;
output int  REQUEST_SEND;    // sends int

// Input events
input int  TEMPERATURE_DONE; // receives int
input void SEND_DONE;

// System calls
function int getRadioID (void);
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
// Output events
void VM.out(int evt_id, void* args) {
    switch (id){
        case O_REQUEST_TEMPERATURE:
            call TINYOS_TEMP.read();
        <...>; // O_REQUEST_SEND
    }
}

// Input events
event TINYOS_TEMP.done (int val) {
    VM.enqueue(I_TEMPERATURE_DONE, &val);
}
<...> // TINYOS_SEND.done

// System calls
void VM.function(int id, void* params) {
    switch (id) {
        case F_GET_RADIO_ID:
            VM.push(TINYOS_NODE_ID);
    }
}
\end{lstlisting}
\end{minipage}
%
\caption{
%{\small %\textmd{
\CEU interface (in the left) with associated customized \VM (in the right).
%
\code{VM.out} redirects all output events to the corresponding OS call (ln.  
1--8).
%
Each \emph{TinyOS} event callback calls \code{VM.enqueue} for the corresponding 
input event (ln 10--14).
%
System calls use \code{VM.push} for immediate return values (ln. 16--22).
%
\label{lst.terra.defs}
}
\end{figure}

Considering the compilation process illustrated in Figure~\ref{fig.impl} for 
the \C back end, the main difference resides in the \emph{Code Generation} 
phase, which here outputs assembly instructions for the Terra \VM, instead of 
\C.
%
The \VM includes special instructions to support complex and recurrent 
operations from the runtime of \CEU, such as dispatching events, clearing 
aborted trails
Basically, they map

Basically, it is composed by instructions to define a pending trail (\code{tkins}),
a pending event(using \code{set\_c} to set a value in the memory),
a check operation to verify the \code{par/and} termination(\code{chkret}),
a pending timer event (\code{clken}), and some others like to clear trails data or suspend execution.
All pending entries are managed by the scheduler, that also responds for timer and input events.
All trails are executed as a TinyOS task, this ensure that each trail runs to completion avoiding race condition between then.

During compilation process the script is broken in trails in the same way it is 
done in the \CEU compiler explained in Section 4.
The Terra virtual machine implements a similar \CEU scheduler to execute the 
script trails, but the VM
uses the real trails address instead of labels. In this way is possible to jump directly to each script
partitions without using a dispatcher.

@@

\begin{comment}
%-------------------------------------------------
\end{comment}

@@

@@

%\emph{(As subsecoes serao retiradas do texto ao final...)}

%\subsection{Semantics}

@@

Many synchronous languages have been designed to interoperate with C, such as 
\emph{Reactive C}~\cite{rp.rc}, Protothreads~\cite{wsn.protothreads}, 
\emph{PRET-C}~\cite{rp.pretc} and \emph{SC}~\cite{rp.synchc}.
%
They offer Esterel-like parallel compositions with communication via shared 
variables, relying on deterministic scheduling to preserve determinism.
%
However, it is the responsibility of the programmer to specify the execution 
order for threads, based on either explicit priorities, or source code lexical 
order.
%
These languages have a tick-based notion of time similar to Esterel, which 
prevents the event-based temporal analysis of \CEU.

@@

%\subsection{Implementation}

@@

%\subsection{VM}

A number of virtual machines have been proposed for embedded systems.
%
The \emph{Sun SPOT} Java-based platform with the \emph{Squawk 
JVM}~\cite{wsn.sunspot} brings Java for the embedded domain, but requires a 
much powerful hardware.%
\footnote{The \emph{Sun SPOT} uses a 32-bit CPU with 4 Mbytes of FLASH and 512 
KBytes of SRAM.}
%
\emph{Darjeeling}~\cite{wsn.darjeeling} and \emph{TakaTuka}~\cite{wsn.takatuka} 
are complete \emph{JVMs} targeting constrained embedded systems with support 
for multithreading and garbage collection.
%
Java has antagonistic design choices in comparison to \CEU:
it does not impose static bounds on memory usage and execution time, and 
provides preemptive multithreading with synchronization primitives for 
accessing shared memory.
%
Plummer et al.~\cite{esterel.vm} propose a Esterel-based \VM with similar 
design choices to our work.
To reduce code size, the \VM has a instruction set to deal with event handling 
and concurrency constructs that are particular to Esterel.
However, the proposed \VM is only a proof of concept, with no support for 
arithmetic operations, external system calls, or remote reprogramming.
